Appendix: A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-09-06 JCGS_response_to_reviews/response_to_review.Rmd

# Reviewer 1

*Comments to the Author*

*I did not need any convincing that residual diagnostic plots are superior to tests but this paper presents a narrower argument to support this claim than many analysts would employ. For example, diagnostic plots may reveal features which are not directly connected to a linear model assumption and yet are interesting and important to identify. Of course, it is difficult to justify a journal article on the basis of such observations. In this respect, I am impressed that you have succeeded in demonstrating the superiority within the more formal framework of the lineup protocol. You have restricted attention to the residual-fitted plot and a particular selection of formal tests. I accept that this was necessary to make definite progress. Nevertheless, it would be good to add some discussion of the wider picture.*

Thank for this feedback. We are happy to see that, like many people, you don't need convincing and that you are glad to see that we have succeeded in demonstrating the superiority of residual diagnostic plots.

We have added more explanation about the use of residual plots in the Introduction (highlighted in red) and we have added a new section "Limitations and practicality" that provides some insight into the wider picture.

<!-- (diagnostic plots may reveal features which are not directly connected to a linear model assumption; You have restricted attention to the residual-fitted plot and a particular selection of formal tests; Nevertheless, it would be good to add some discussion of the wider picture.)-->

@@

*1. "The ubiquity of this advice is curious: investigating why, is the subject of this paper. " I don't find this advice curious - it seems almost universal. I have taught this topic for many years and give the same advice. I explain why I give this advice and I believe several other authors do the same. Their reasons are wider in scope than you express in this article.*

We have changed the language.

*2. All your examples are simulated data and very obviously so. Real data shows much greater irregularity - the points will usually not be uniformly spaced along the x-axis. The response may be somewhat discrete or have other features that will distinguish it from bland randomness. This would seem to offer greater challenges to your lineup idea as generating data under the null will be much more challenging. If we simulate from the null model, the real data will be very obvious. You have done an experiment but the reality would be more like an observational study. This limits the scope of your conclusions which you should clearly acknowledge.*

The work very specifically uses simulated data because we need to study a problem in controlled conditions. You are right, though, that data will pose more interesting challenges, and this can be in two directions. One, being that the lineup protocol will allow interesting and unexpected patterns in the residuals to be detected (see Buja et al and Wickham et al for examples). The other is that the current process for generating null data uses residual rotation and will likely maintain irrelevant quirks of the data in the null plots. These should not inadvertently allow the reader to detect the data plot based on quirks.

We have addressed this criticism in the broader discussion of the use of residual plots and lineups for residual diagnostics added to the new Section "Limitations and practicality". <!-- (see issue #11).-->

An observational study involving actual residual plots will hopefully be possible once a computer vision model is deployed that will allow analysts to upload their residuals for automated testing.

@@

*3. It is true that formal diagnostic tests tend to be too sensitive and yet we don't have a definite sense of how much divergence from the assumed model would cause a significant problem. How much non-normality in the residuals is acceptable? We don't really know. Formal tests give us definite answers about violations of assumptions which we might find hard to ignore but diagnostic plots give us license to sweep inconvenient features under the carpet. Is there really an advantage to the diagnostic plot in this case other than the opportunity to hand-wave about our answer?*

When the diagnostic plots are embedded in a lineup of nulls, it does have the advantage of providing a calibrated answer, and protects against the subjectiveness of hand-waving. We are advocating that you should always calibrate the diagnostic plot with a comparison with null plots. We have strengthened this point of view by adding more explanation in the Introduction, in addition to what is provided in the Conclusions.

Having a definitive (numerical) answer to "how much is too much" would require more and different work. For our purposes, it was that the true residual plot could be detected from the null plots when the lineup was evaluated by K people. We've added a sentence to the Conclusion mentioning this.

<!--(Is there a paper to refer the reader to?)-->

@@

*4. I use something like the lineup protocol in teaching to help students calibrate their assessment of diagnostic plots. But I doubt very much experienced analysts use them in practice. I do not think it is realistic or really that helpful to recommend their everyday use.*

It can be argued that it is being used by practitioners based on download rates of the nullabor package, and applications articles that report usage. Using it formally, we agree is cumbersome and would require engaging a crowd-sourcing service, that involves a small expense. Science labs typically have a budget for equipment and experimental devices and we would advocate this is a useful addition. We have definitely used it ourselves in 10.1007/s00180-014-0534-x for ecological data and  10.4172/2153-0602.1000139 for gene expression analysis. We have added some details in the "Limitations and practicality" section. <!--We agree that it is not widespread, and think that this is because it's a little unwieldy. The goal is to make it easier, particularly by employing an automated testing tool using computer vision, but this is not the subject of the current paper.-->

@@

*This article uses the lineup protocol in a visual inference experiment, which puts a residual plot in the context of null plots. Focus is mostly on detecting nonlinearities and heteroskedasticity. Comparisons with some classic tests are given.*

*Overall Comments*

*Overall, this paper is very well written and interesting. However, I have some serious reservations about the practicality of the procedure, which I articulate in greater detail below. I also think the paper is currently too long, and that a significant chunk of material should be shifted to the Appendix.*

Thanks for the positive feedback. We have shortened the paper by XXX pages. We have expanded the explanations on the practicality in the Section "Limitations and practicality".

@@

*1. As mentioned above, the paper is too long. While the onus should be on the authors for getting the page number down, some possible suggestions could be combining Sections 1 and 2 while moving some of the historical context (which I fully acknowledge is quite interesting) to the Appendix, and moving some of the technical details of Section 3.2 to the Appendix.*

Thanks for your suggestions. This is how we have shortened the paper:

- Details of Section 3.2 Statistical significance moved to the Appendix. 
- Sections 1 and 2 have been made more precise, while still covering the requested additional topics.
- The explanation of power calculation was left in the main paper because this is new and of primary importance for understanding the results.

- Part of Experimental design -> appendix

@@

*2. The practical implementation of this as a procedure is, in my opinion, the biggest weakness. The authors acknowledge the costliness of human evaluation of residuals, but not until Page 33, Line 47. I find this a tad disingenuous as this will likely jump out to the reader as being a problem right from the beginning of reading this paper (it did for me). So I think being upfront about this limitation, and providing some practical considerations to address this concern will assuage my concerns. There might be some concepts from human-computer interaction that could be effective for operationalizing and improving the efficiency of the present work.*

The new Section "Limitations and practicality" addresses these concerns and we have added two sentences to the Introduction.

@@

*3. I think a better connection needs to be made with where this computationally expensive process could potentially pay off. For example, applying the lineup protocol to, say, any of the data examples in the texts cited by the authors would amount to nothing more than an academic exercise. But, what sort of data problems can the authors point to and definitively claim that the lineup protocol will be invaluable at ensuring we visually capture possible nonlinearities or heteroskedasticity? I am not doubting the efficacy of the approach, but I’d like to see a stronger case made with its practical application.*

See the new Section "Limitations and practicality". Some additional text is provided in various places. It's very practical to employ the lineup protocol using the `nullabor` package, and the (growing) number of citations from publications in a variety of areas show that it is being used.

<!-- As mentioned above,

This is not computationally expensive. The conventional test also make mistakes. You need something more than the conventional test. And you also need something more than the residual plot.

check earlier paper.-->

@@

*4. What should be claimed about the sample size n as a limitation? Take in point the comments at the bottom of page 19. Am I to understand that the lineup protocol will not be appropriate for big data problems?*

The lineup protocol is very suitable for big data. A plot other than a scatterplot is probably needed to better display a large number of observations. We have changed the language so that this is clearer.

@@

*5. At the beginning of Section 4.2.3, what cultural biases do the authors anticipate by restricting the participants to be “fluent in English”? Granted this is likely done as a matter of convenience, but it would be good to articulate how cultural biases may be present in interpreting a lineup protocol.*

The instructions for this study were in English, which motivated the constraint applied in the crowd-sourcing service. However many participants were likely multi-lingual, and they joined from a variety of countries. We don't anticipate that the results would be different with a different sample.

@@

*6. I would like the authors to be a little clearer about their specific contributions. I felt as if the biggest contribution in this work was the experiments that they did. A lot of references are included for work that is applied in this manuscript, so at times I got lost as to what already existed in the literature and what “new” material the authors were giving the reader.*

We have clarified the language of the Introduction and the Conclusion.

*Minor Comments*

*1. Page 9, Line 13: Since the three references are by the same authors, use \\cite{Ref1,Ref2,Ref3}.* DONE

*2. Page 12, Line 7: $\alpha$.* DONE

*3. Why not write the power formula in terms of a generic significance level? Granted, one should be mindful of an abuse of notation if using $\alpha$ as it has been introduced in the Rorschach setup of the previous subsection.* We have kept the current explanation of power.

*4. Page 13, Line 14: "...the scenario..."* DONE

*5. Page 20, Line 36: "...variety of difficulty..."* DONE

@@

*6. In Figures 11 and 12, I am not sure the example residual plots beneath the x-axis are all that terribly helpful. In fact, I think they all look mostly similar, and am not entirely convinced that they add any value. Can the authors convince me otherwise?*

The small residual plots help to understand the change in effect size represented by the numerical values on the x-axis. They are indispensable, as reported by numerous readers during the proofing stage of the paper.

### 2023-09-06 appendix.Rmd
double space
finished revising the paper

---
title: |
  Appendix: A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

# Additional details of testing procedures
## Effect size derivation

>>>>

---
title: |
  Appendix: A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

# Additional details of testing procedures

## Statistical significance

Within the context of visual inference, with $K$ independent observers, the visual $p$-value can be seen as the probability of having as many or more participants detect the data plot than the observed result.

The approach used in @majumder2013validation is as follows. Define $X_j = \{0,1\}$ to be a Bernoulli random variable measuring whether participant $j$ detected the data plot, and $X = \sum_{j=1}^{K}X_j$ be the total number of observers who detected the data plot. Then, by imposing a relatively strong assumption that all $K$ evaluations are fully independent, under $H_0$ $X \sim \mathrm{Binom}_{K,1/m}$. Therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is estimated with $P(X \geq x) = 1 - F(x) + f(x)$, where $F(.)$ is the binomial cumulative distribution function, $f(.)$ is the binomial probability mass function and $x$ is the realization of number of observers choosing the data plot.

As pointed out by @vanderplas2021statistical, this basic binomial model is deficient. It does not take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup, or account for when participants are offered the option to select one or more "most different" plots, or none, from a lineup. They suggest three common lineup scenarios: (1) $K$ different lineups are shown to $K$ participants, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ participants, and (3) the same lineup is shown to $K$ participants. Scenario 3 is the most feasible to apply, but has the most dependencies to accommodate for the $p$-value calculation. For Scenario 3, VanderPlas et al propose modelling the probability of plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim \text{Dirichlet}(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case participant $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots participant $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual $p$-value is given as

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}
```

\noindent where $B(.)$ is the beta function defined as

```{=tex}
\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}
```

\noindent We extend the equation to non-negative real number $c_i$ by applying a linear approximation

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}
```

\noindent where $P(C \geq \lceil c_i \rceil)$ is calculated using Equation \ref{eq:pvalue-beta-binomial} and $P(C = \lfloor c_i \rfloor)$ is calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

The parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and will need to be estimated from the survey data. An interpretation of $\alpha$ is that when it is low only a few plots are attractive to the observers and tend to be selected, and when high, most plots are equally likely to be chosen. VanderPlas et al define $c$-interesting plot to be if $c$ or more participants select the plot as the most different. The expected number of plots selected at least $c$ times, $\mathrm{E}[Z_c]$, is then calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
\mathrm{E}[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. Precise estimation of $\alpha$, is aided by evaluation of Rorschach lineups, where all plots are null plots. In a Rorschach, in theory all plots should be equally likely, but in practice some (irrelevant) visual elements may be more eye-catching than others. This is what $alpha$ captures, the capacity for extraneous features to distract the observer for a particular type of plot display.

## Effect size derivation

### 2023-09-06 appendix.tex
double space
finished revising the paper

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}

@@

>>>>

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\doublespacing

@@

\hypertarget{statistical-significance}{%
\subsection{Statistical significance}\label{statistical-significance}}

Within the context of visual inference, with \(K\) independent
observers, the visual \(p\)-value can be seen as the probability of
having as many or more participants detect the data plot than the
observed result.

The approach used in \citet{majumder2013validation} is as follows.
Define \(X_j = \{0,1\}\) to be a Bernoulli random variable measuring
whether participant \(j\) detected the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the total number of observers who detected
the data plot. Then, by imposing a relatively strong assumption that all
\(K\) evaluations are fully independent, under \(H_0\)
\(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value of a
lineup of size \(m\) evaluated by \(K\) observer is estimated with
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the binomial
cumulative distribution function, \(f(.)\) is the binomial probability
mass function and \(x\) is the realization of number of observers
choosing the data plot.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model is deficient. It does not take into account the possible
dependencies in the visual test due to repeated evaluations of the same
lineup, or account for when participants are offered the option to
select one or more ``most different'' plots, or none, from a lineup.
They suggest three common lineup scenarios: (1) \(K\) different lineups
are shown to \(K\) participants, (2) \(K\) lineups with different null
plots but the same data plot are shown to \(K\) participants, and (3)
the same lineup is shown to \(K\) participants. Scenario 3 is the most
feasible to apply, but has the most dependencies to accommodate for the
\(p\)-value calculation. For Scenario 3, VanderPlas et al propose
modelling the probability of plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim \text{Dirichlet}(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case participant
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots participant \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
\(p\)-value is given as

\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}

\noindent where \(B(.)\) is the beta function defined as

\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}

\noindent We extend the equation to non-negative real number \(c_i\) by
applying a linear approximation

\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}

\noindent where \(P(C \geq \lceil c_i \rceil)\) is calculated using
Equation \ref{eq:pvalue-beta-binomial} and
\(P(C = \lfloor c_i \rfloor)\) is calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

The parameter \(\alpha\) used in Equation \ref{eq:pvalue-beta-binomial}
and \ref{eq:pmf-beta-binomial} is usually unknown and will need to be
estimated from the survey data. An interpretation of \(\alpha\) is that
when it is low only a few plots are attractive to the observers and tend
to be selected, and when high, most plots are equally likely to be
chosen. VanderPlas et al define \(c\)-interesting plot to be if \(c\) or
more participants select the plot as the most different. The expected
number of plots selected at least \(c\) times, \(\mathrm{E}[Z_c]\), is
then calculated as

\begin{equation} \label{eq:c-interesting-expectation}
\mathrm{E}[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}

With Equation \ref{eq:c-interesting-expectation}, \(\alpha\) can be
estimated using maximum likelihood estimation. Precise estimation of
\(\alpha\), is aided by evaluation of Rorschach lineups, where all plots
are null plots. In a Rorschach, in theory all plots should be equally
likely, but in practice some (irrelevant) visual elements may be more
eye-catching than others. This is what \(alpha\) captures, the capacity
for extraneous features to distract the observer for a particular type
of plot display.
